\chapter{Mathematical Probability Models}
\section{Sample Spaces}
A sample space,~\sS, is a set of distinct outcomes for an experiment or process, with the property that in a single trial, one~and only~one of these outcomes occurs. The outcomes that make up a sample space are called sample points or simply points.
\begin{example}
The sample space for a roll of a six-sided die is
\[
    \{a_1,a_2,a_3,a_4,a_5,a_6\}\hspace{5mm}\text{where $a_i$ is the event the top face is $i$}
\]
More simply we could define the sample space as
\[
    \{1,2,3,4,5,6\}
\]
\end{example}
\begin{info}
Note that a sample space of a probability model for a process is not necessarily unique. Often times, however, we try to chose sample points that are the smallest possible or ``indivisible''.
\end{info}
\begin{example}
If we define $E$ to be the event that the top face of a six-sided~die is even when rolled and $O$ to be the event the top-face is odd, then the sample space,~\sS, can be defined as
\[
    \{E,O\}
\]
This is the same process as Example 2.1.1 (rolling a six-sided die), so since the sample spaces differ, clearly, sample spaces are not unique. Moreover, if we are interested in the event that a 3 is rolled, this sample space is not suitable since it groups the event in question with other events.
\end{example}
A sample space can be either \textbf{discrete} or \textbf{non-discrete}. If a sample space is discrete, it consists of a finite or countably infinite number ``simple~events''. A countably infinite set is one that can be put into a one-to-one correspondence with the set of real numbers. For example, $\left\{1,\frac{1}{2},\frac{1}{3},\frac{1}{4},\ldots\right\}$ is countably infinite whereas $\{\,x \ssep x\in\sR\,\}$ is not.
\subsection*{Simple Events}
An event in a discrete sample space is a subset of the sample space, i.e., $A \subset \sS$. If the event is indivisible, so as to only contain one point, we call it a simple event, otherwise it is a compound event.
\begin{example}
A simple event for a roll of a six-sided die is $A = \{a_1\}$ where $a_i$ is the event the top face is i. A compound event is $E = \{a_2,a_4,a_6\}$.
\end{example}

\section{Assigning Probabilities}
Let $\sS = \{a_1,a_2,a_3,\ldots\}$ be a discrete sample space. We assign probabilities, $P(a_i),\for i = 1,2,3,\ldots$ to each sample point $a_i$ such that the following two conditions hold
\begin{itemize}
    \item $0 \leq P(a_i) \leq 1$
    \item $\sum\limits_{\all i} P(a_i) = 1$
\end{itemize}
The set of probabilities $\{\,P(a_i) \ssep i = 1,2,3,\ldots\,\}$ is called a \textbf{probability distribution} on \sS.
\begin{info}
Note that $P$ is a function with the sample space as its domain.
\end{info}
The second condition, that the sum of the probabilities of all sample points is 1, relates to the property that for a given experiment one simple event in the sample space must occur. Every experiment or process always has an outcome thus the probability of any outcome being achieved must be 1.
\subsection*{Compound Events}
The probability of an event $A$ is the sum of the probability of all the simple events that make up $A$.
\[
    P(A) = \sum_{a \in A}P(a)
\]
\begin{example}
In the previous example we saw that $E = \{a_2,a_4,a_6\}$ is a compound event. Thus, the probability of the compound event $E$ is
\[
    P(E) = P(a_2) + P(a_4) + P(a_6)
\]
\end{example}
\begin{info}
Note that the probability model that we defined does not specify what actual numbers to assign to the simple events of a process. It only defines the properties that guarantee mathematical consistency. Thus, if we assigned $P(a_2)$ to be $0.9$, our model would still be mathematically consistent but would not be consistent with the frequencies we obtain in multiple repetitions of the experiment.
\end{info}
\begin{info}
In actual practice we try to define probabilities that are approximately consistent with the frequencies of the events in multiple repetition of the process.
\end{info}
\subsection*{Complements}
The complement of an event, $A$, is the set of all outcomes not included in $A$ and is denoted by $\comp{A}$.
\begin{example}
If $E = \{\, a_1, a_3, a_5 \,\}$ is a compound event on the sample space $\{\, a_1, a_2, a_3,a_4,a_5,a_6 \,\}$, then the complement of $E$ is
\[
    \comp{E}=\{\, a_2,a_4,a_6 \,\}
\]
\end{example}
\begin{info}
Because of the nature of complementary events, two complementary events cannot both occur in one process. The events are \textbf{mutually exclusive}.
\end{info}
